\input{preamble}

\newif \ifDraft\Drafttrue
%\Draftfalse

\ifDraft
  \newcommand{\Comment}[1]{\textbf{\textsl{#1}}}
\else
  \newcommand{\Comment}[1]{\relax}
\fi

\newcommand{\gernot}[1]{\Comment{#1 [gernot]}}


\title{\vspace{-0.8cm}\textbf{
Research Proposal: Exploiting Heterogeneity in Modern Computer Architectures
}}

\author{
Natacha Crooks$^1$ \quad
Micheal Walter-Echols$^2$ \quad
Manohar Vanga$^1$
\\ \emph{$^1$ MPI-SWS, $^2$ Saarland University}
}
\date{\vspace{-12pt}}


% for the draft, please use double spacing 
%\doublespacing

\begin{document}
\maketitle

% for camera-ready versions
%\thispagestyle{empty}
%\pagestyle{empty}

%\linenumbers
%\setlength\linenumbersep{0.1cm}


\input{abs}

\section{We should probably define what
we mean by heterogeinity}

\begin{itemize}
\item In~\cite{SPBRBHI:08}, define three types of heterogeneity:
\begin{itemize}
\item non-uniformity.  This refers both to memory (NUMA), but also
to communication cost within a processor
\item core diversity. Cores will have different performance, possibly
even different ISAs.
\item system diversity. Different machines will be very diverse.
\end{itemize}
\item We are specifically interested in the first two, so heterogeineity
within a single machine.
\item Non-uniformity:
\item Is heterogeneity a necessary evil that has arisen as
a result of the inability to get more performance out of single-core
and the increasigly stringent power requirements. Ak,
are current systems suffering from heterogeinity. (for ex,
in distributed systems, the use of commodity hardware is a necessary evil)
\item Or is it something which should be leveraged, and we are currently
not doing so sufficiently?
\item What are the drawbacks of current approaches?
\end{itemize}
\section{brief plan of action}
\begin{itemize}
 \item understand what the tradeoffs are
when building a heterogeneous systems
 \item come up with very cool idea
\end{itemize}
%\scriptsize
\fontsize{8}{8.2}
\selectfont
\bibliographystyle{plain}
\bibliography{main}
\normalsize

\end{document}             % End of document.
